---
title: "projectwork"
author: "Amanda Chang"
date: "2026-04-03"
output: html_document
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)     # for data cleaning and plotting
library(DT)            # for visualizing data tables 
library(janitor)       # for automatically renaming variables
library(ggmap)         # for mapping points on maps
library(RColorBrewer)  # for color palettes
library(sf)            # for working with spatial data
library(leaflet)       # for highly customizable mapping
library(ggthemes)      # for more themes (including theme_map())
library(gplots)        # for col2hex() function
library(choroplethr)   # another option to make choropleths
library(choroplethrMaps)
library(maps)
library(leaflet.extras)
library(htmlwidgets) # using this for my new legend
```

```{r, include=FALSE, echo=FALSE}
boston_crashes_raw <- read.csv("visionzerocrash.csv")

bike_network <- st_read("existing_bike_network_2024.geojson") %>% 
  st_transform(4326) # Ensure WGS84 for Leaflet
head(bike_network)
```

```{r, echo=FALSE}
# sorted from most dangerous (top of legend) to safest (bottom of legend)
safety_order <- c(
  "SLM",        # Shared Lane Markings (Zero protection, full mixing)
  "BL-PEAKBUS", # Bike Lane/Peak Hour Bus Lane (High-speed heavy vehicle mixing)
  "CFBS",       # Contraflow Bike Lane (typo found in the database)
  "BLSL",       # Bike Lane/Shared Lane Hybrid (Inconsistent protection)
  "SBLSL",      # Separated Bike Lane/Shared Lane Hybrid (Gaps in protection)
  "SLMTC",      # Traffic-Calmed Local Street with Shared Lane Markings
  "CFBL",       # Contraflow Bike Lane (Painted-only, high conflict risk)
  "BL",         # Bike Lane (Standard painted lane, door-zone risk)
  "BFBL",       # Buffered Bike Lane (Paint buffer, but no physical barrier)
  "SBLBL",      # Separated Bike Lane/Bike Lane Hybrid
  "SBL",        # Separated Bike Lane (Standard flex-post or curb)
  "CFSBL",      # Contraflow Separated Bike Lane (High protection)
  "SUPN",       # Shared Use Path - Natural Surface (Safe from cars, but slip risk)
  "SUPM",       # Shared Use Path - Minor
  "SUP",        # Shared Use Path (Fully off-street)
  "WALK",       # Walkway
  "PED"         # Pedestrian Zone (Lowest vehicle interaction)
)

bike_labels <- data.frame(
  Abbr = safety_order,
  infra_full_name = c(
    "Shared Lane Markings",
    "Bike Lane/Peak Hour Bus Lane",
    "Contraflow Bike Lane",
    "Bike Lane Side + Shared Lane Side",
    "Separated Bike Lane Side + Shared Lane Side",
    "Traffic-Calmed Local Street",
    "Contraflow Bike Lane",
    "Standard Bike Lane",
    "Buffered Bike Lane",
    "Separated/Standard Hybrid",
    "Separated Bike Lane",
    "Contraflow Separated Bike Lane",
    "Shared Use Path (Natural)",
    "Shared Use Path (Minor)",
    "Shared Use Path (Paved)",
    "Walkway",
    "Pedestrianized Street"
  )
)

exposure_list <- data.frame(safety_order)
exposure_list$exposure_metric <- 1:nrow(exposure_list)
```


```{r, echo=FALSE}
bike_crashes <- boston_crashes_raw %>%
   filter(mode_type == "bike") %>%
   drop_na(lat, long) %>%
   st_as_sf(coords = c("long", "lat"), crs = 4326)

roads_projected <- st_transform(bike_network, 2249)
crashes_projected <- st_transform(bike_crashes, 2249)

nearest_indices <- st_nearest_feature(crashes_projected, roads_projected)
nearest_distances <- st_distance(
  crashes_projected,
  roads_projected[nearest_indices, ],
  by_element = TRUE
)

# if its within 15 meters then keep it
threshold_m  <- 15

assigned_indices <- ifelse(
  as.numeric(nearest_distances) <= threshold_m,
  nearest_indices,
  NA  # dont assign roads too far from the road
)

road_counts <- table(assigned_indices[!is.na(assigned_indices)])

bike_network$crash_count <- 0
bike_network$crash_count[as.numeric(names(road_counts))] <- as.numeric(road_counts)
bike_network$ExisFacil <- factor(bike_network$ExisFacil, levels = safety_order, ordered=TRUE)
bike_network <- bike_network %>%
  left_join(bike_labels, by = c("ExisFacil" = "Abbr"))


danger_colors <- colorRampPalette(c("#e41a1c", "#9B8FFF", "#00A719"))(length(safety_order))
names(danger_colors) <- safety_order

pal <- colorFactor(
  palette = danger_colors,
  levels = safety_order,
  domain = bike_network$ExisFacil
)


leaflet() %>%
   addProviderTiles(providers$CartoDB.Positron) %>%

   addPolylines(
      data = bike_network,
      color = ~pal(ExisFacil),
      weight = 3,
      opacity = 1,
      label = ~paste0("Street: ", STREET_NAM, " | Crashes: ", crash_count," | Infra: ", infra_full_name),
      group = "Infrastructure"
   ) %>%
  
  addHeatmap(
    data = bike_crashes,
    lng = ~st_coordinates(bike_crashes)[,1],
    lat = ~st_coordinates(bike_crashes)[,2],
    gradient = c(
    '0.4' = 'darkblue',
    '0.6' = 'blue',
    '0.7' = 'yellow', 
    '0.8' = 'orange', 
    '1.0' = 'red'
    ),
    blur = 15,
    max = 0.11,
    radius = 11,
    group = "Crash Hotspots"
  ) %>%
  
  addControl(
  html = "<div id='infra-legend' style='background: rgba(255, 255, 255, 0.5); 
                      padding: 10px; 
                      border: 3px solid rgba(0,0,0,0.2); 
                      border-radius: 4px; 
                      text-align: center;
                      font-family: sans-serif;'>
            
            <b style='display: block; margin-bottom: 8px; font-size: 12px;'>Vehicle Exposure</b>
            
            <div style='font-size: 10px; font-weight: bold; color: #e41a1c; margin-bottom: 2px;'>MOST</div>
            
            <div style='height: 120px; 
                        width: 18px; 
                        background: linear-gradient(to bottom, #e41a1c 0%, #6250FF 50%, #00A719 100%); 
                        margin: 2px auto;
                        border-radius: 2px;'>
            </div>
            
            <div style='font-size: 10px; font-weight: bold; color: #00A719; margin-top: 2px;'>LEAST</div>
            
          </div>",
  position = "bottomright"
  ) %>% 

  addLayersControl(
    overlayGroups = c("Infrastructure", "Crash Hotspots"),
    options = layersControlOptions(collapsed = FALSE)
  ) %>%
  
  onRender("
    function(el, x) {
      var map = this;
      var legend = document.getElementById('infra-legend');
      
      map.on('overlayadd', function(e) {
        if (e.name === 'Infrastructure') {
          legend.style.display = 'block';
        }
      });
      
      map.on('overlayremove', function(e) {
        if (e.name === 'Infrastructure') {
          legend.style.display = 'none';
        }
      });
    }
  ")
```


```{r, echo=FALSE}
dangerous_roads_stats <- bike_network %>%
  as_tibble() %>% 
  group_by(STREET_NAM) %>%
  summarize(total_crashes = sum(crash_count, na.rm = TRUE)) %>%
  filter(total_crashes > 20) %>% 
  arrange(desc(total_crashes))

ggplot(dangerous_roads_stats, aes(x = reorder(STREET_NAM, total_crashes), y = total_crashes)) +
  geom_col(fill = "#b90e0a") +
  coord_flip() +
  labs(
    title = "Total Bike Crashes Ranked by Road",
    subtitle = "Roads with over 20 crashes",
    x = "Street Name",
    y = "Number of Crashes"
  )

```


According to [this article by Boston Magazine ](https://www.bostonmagazine.com/news/2017/03/01/boston-bike-traffic-data/#:~:text=Massachusetts%20Ave.,of%20Columbus%20Avenue%20%E2%80%94%20797%20bicyclists) which recorded the busiest cycling roads in Boston, the top ten Boston roads with the most cyclists are

1. Massachusetts Ave. Bridge, south of Back Street — 3,081 bicyclists

2. BU Bridge, north of Commonwealth Avenue — 1,957 bicyclists

3. Commonwealth Avenue, west of Silber Way — 1,571 bicyclists

4. Southwest Corridor Bicycle Path, south of Heath Street — 1,554 bicyclists

5. Longwood Avenue, east of Pilgrim Road — 1,348 bicyclists

6. North Harvard Street, south of Soldiers Field Road — 1,320 bicyclists

7. Columbus Avenue, west of Massachusetts Avenue — 1,314 bicyclists

8. Brighton Avenue, east of St. Lukes Road — 1,063 bicyclists

9. Columbus Avenue, west of Holyoke Street — 825 bicyclists

10. Massachusetts Avenue, south of Columbus Avenue — 797 bicyclists

```{r, echo=FALSE}
busy_street_names <- c(
  "Massachusetts Avenue", "Commonwealth Avenue", "Columbus Avenue", 
  "Brighton Avenue", "North Harvard Street", "Longwood Avenue", 
  "Boston University Bridge", "Southwest Corridor"
)
busy_road_stats <- bike_network %>%
  as_tibble() %>% 
  filter(STREET_NAM %in% busy_street_names) %>%
  group_by(STREET_NAM) %>%
  summarize(total_crashes = sum(crash_count, na.rm = TRUE)) %>%
  arrange(desc(total_crashes))

ggplot(busy_road_stats, aes(x = reorder(STREET_NAM, total_crashes), y = total_crashes)) +
  geom_col(fill = "#b90e0a") +
  coord_flip() +
  labs(
    title = "Total Bike Crashes on Boston's Busiest Cycling Roads",
    subtitle = "Based on Vision Zero Crash Data and Boston Magazine's Top 10 List",
    x = "Street Name",
    y = "Number of Crashes"
  ) +
  theme_minimal()
```



```{r, echo=FALSE}
traffic_counts <- tibble(
  STREET_NAM = c("Massachusetts Avenue", "Boston University Bridge", "Commonwealth Avenue", 
                 "Southwest Corridor", "Longwood Avenue", "North Harvard Street", 
                 "Columbus Avenue", "Brighton Avenue"),
  daily_bikers = c(3878, 1957, 1571, 1554, 1348, 1320, 2139, 1063)
)

normalized_stats <- bike_network %>%
  as_tibble() %>%
  filter(STREET_NAM %in% traffic_counts$STREET_NAM) %>%
  group_by(STREET_NAM, ExisFacil) %>%
  summarize(total_crashes = sum(crash_count, na.rm = TRUE), .groups = "drop") %>%
  left_join(traffic_counts, by = "STREET_NAM") %>%
  mutate(crash_rate = (total_crashes / daily_bikers) * 100)

ggplot(normalized_stats, aes(x = reorder(STREET_NAM, crash_rate, sum), 
                             y = crash_rate)) +
  geom_col(fill = "#b90e0a") +
  coord_flip() +
  scale_fill_viridis_d() +
  labs(
    title = "Bike Crash Rate on Boston's Busiest Cycling Roads",
    subtitle = "Normalized: (Total Crashes / Daily Bicyclists) * 100",
    x = "Street Name",
    y = "Total Crashes Recorded / Daily Riders"
  ) +
  theme_minimal()
```

# ```{r, fig.width=10, fig.height=6, dpi=300}
# normalized_stats <- normalized_stats %>%
#   mutate(ExisFacil = fct_recode(ExisFacil,
#     "Shared Lane Markings" = "SLM",
#     "Peak Hour Bus/Bike Lane" = "BL-PEAKBUS",
#     "Contraflow Bike Lane" = "CFBS", # bike street is not real
#     "Hybrid Shared Lane" = "BLSL",
#     "Separated/Shared Hybrid" = "SBLSL",
#     "Traffic-Calmed Shared Lane" = "SLMTC",
#     "Contraflow Bike Lane" = "CFBL",
#     "Standard Bike Lane" = "BL",
#     "Buffered Bike Lane" = "BFBL",
#     "Separated/Standard Hybrid" = "SBLBL",
#     "Separated Bike Lane" = "SBL",
#     "Contraflow Separated Lane" = "CFSBL",
#     "Natural Shared Use Path" = "SUPN",
#     "Minor Shared Use Path" = "SUPM",
#     "Shared Use Path" = "SUP",
#     "Walkway" = "WALK",
#     "Pedestrian Zone" = "PED"
#   ))
# 
# ggplot(normalized_stats, aes(x = reorder(STREET_NAM, crash_rate, sum), 
#                              y = crash_rate, 
#                              fill = ExisFacil)) +
#   geom_col() +
#   coord_flip() +
#   scale_fill_viridis_d() + 
#   labs(
#     title = "Bike Crash Rate on Busiest Boston Roads",
#     subtitle = "Normalized by Number of Daily Bicyclists (Boston Magazine)",
#     x = "Street Name",
#     y = "Total Crashes per 100 Daily Bikers",
#     fill = "Bike Infrastructure Type"
#   ) +
#   theme_minimal() +
#   theme(legend.position = "bottom",
#         legend.text = element_text(size = 8)) +
#   guides(fill = guide_legend(ncol = 4))
# ```