---
title: "projectwork"
author: "Amanda Chang"
date: "2026-04-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)     # for data cleaning and plotting
library(DT)            # for visualizing data tables 
library(janitor)       # for automatically renaming variables
library(ggmap)         # for mapping points on maps
library(RColorBrewer)  # for color palettes
library(sf)            # for working with spatial data
library(leaflet)       # for highly customizable mapping
library(ggthemes)      # for more themes (including theme_map())
library(gplots)        # for col2hex() function
library(choroplethr)   # another option to make choropleths
library(choroplethrMaps)
library(maps)
library(leaflet.extras)
```

```{r, include=FALSE}
boston_crashes_raw <- read.csv("visionzerocrash.csv")

bike_network <- st_read("existing_bike_network_2024.geojson") %>% 
  st_transform(4326) # Ensure WGS84 for Leaflet
head(bike_network)

bike_network_2022 <- st_read("existing_bike_network_2022.geojson") %>% 
  st_transform(4326) # Ensure WGS84 for Leaflet
head(bike_network_2022)
```

```{r}
unique(bike_network$ExisFacil)
```

```{r}
bike_crashes <- boston_crashes_raw %>%
  filter(mode_type == "bike") %>%
  drop_na(lat, long) %>%
  st_as_sf(coords = c("long", "lat"), crs = 4326)

roads_projected <- st_transform(bike_network, 2249)
crashes_projected <- st_transform(bike_crashes, 2249)

nearest_road_indices <- st_nearest_feature(crashes_projected, roads_projected)

road_counts <- table(nearest_road_indices)
bike_network$crash_count <- 0
bike_network$crash_count[as.numeric(names(road_counts))] <- as.numeric(road_counts)

# sorted from most dangerous (top of legend) to safest (bottom of legend)
safety_order <- c(
  "SLM",        # Shared Lane Markings (Zero protection, full mixing)
  "BL-PEAKBUS", # Bike Lane/Peak Hour Bus Lane (High-speed heavy vehicle mixing)
  "CFBS",       # Contraflow Bike Street (Shared space with opposing traffic)
  "BLSL",       # Bike Lane/Shared Lane Hybrid (Inconsistent protection)
  "SBLSL",      # Separated Bike Lane/Shared Lane Hybrid (Gaps in protection)
  "SLMTC",      # Traffic-Calmed Local Street with Shared Lane Markings
  "CFBL",       # Contraflow Bike Lane (Painted-only, high conflict risk)
  "BL",         # Bike Lane (Standard painted lane, door-zone risk)
  "BFBL",       # Buffered Bike Lane (Paint buffer, but no physical barrier)
  "SBLBL",      # Separated Bike Lane/Bike Lane Hybrid
  "SBL",        # Separated Bike Lane (Standard flex-post or curb)
  "CFSBL",      # Contraflow Separated Bike Lane (High protection)
  "SUPN",       # Shared Use Path - Natural Surface (Safe from cars, but slip risk)
  "SUPM",       # Shared Use Path - Minor
  "SUP",        # Shared Use Path (Fully off-street)
  "WALK",       # Walkway
  "PED"         # Pedestrian Zone (Lowest vehicle interaction)
)

bike_network$ExisFacil <- factor(bike_network$ExisFacil, levels = safety_order)


pal <- colorFactor(
  palette = c("#e41a1c", "#6250FF", "#00A719"),
  domain = bike_network$ExisFacil
)

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%

  addPolylines(
    data = bike_network,
    color = ~pal(ExisFacil),
    weight = 3,
    opacity = 1,
    label = ~paste0("Street: ", STREET_NAM, " | Crashes: ", crash_count),
    group = "Infrastructure"
  ) %>%

  addHeatmap(
    data = bike_crashes,
    lng = ~st_coordinates(bike_crashes)[,1],
    lat = ~st_coordinates(bike_crashes)[,2],
    gradient = c(
    '0.4' = 'darkblue',
    '0.6' = 'blue',
    '0.7' = 'yellow', 
    '0.8' = 'orange', 
    '1.0' = 'red'
    ),
    blur = 15,
    max = 0.11,
    radius = 11,
    group = "Crash Hotspots"
  ) %>%

  addLayersControl(
    overlayGroups = c("Infrastructure", "Crash Hotspots"),
    options = layersControlOptions(collapsed = FALSE)
  ) %>%
  addLegend(
    pal = pal,
    values = bike_network$ExisFacil,
    title = "Road Category",
    position = "bottomright"
  )
```

According to [this article by Boston Magazine ](https://www.bostonmagazine.com/news/2017/03/01/boston-bike-traffic-data/#:~:text=Massachusetts%20Ave.,of%20Columbus%20Avenue%20%E2%80%94%20797%20bicyclists) which recorded the busiest cycling roads in Boston, the top ten Boston roads with the most cyclists are

1. Massachusetts Ave. Bridge, south of Back Street — 3,081 bicyclists

2. BU Bridge, north of Commonwealth Avenue — 1,957 bicyclists

3. Commonwealth Avenue, west of Silber Way — 1,571 bicyclists

4. Southwest Corridor Bicycle Path, south of Heath Street — 1,554 bicyclists

5. Longwood Avenue, east of Pilgrim Road — 1,348 bicyclists

6. North Harvard Street, south of Soldiers Field Road — 1,320 bicyclists

7. Columbus Avenue, west of Massachusetts Avenue — 1,314 bicyclists

8. Brighton Avenue, east of St. Lukes Road — 1,063 bicyclists

9. Columbus Avenue, west of Holyoke Street — 825 bicyclists

10. Massachusetts Avenue, south of Columbus Avenue — 797 bicyclists

```{r}
busy_street_names <- c(
  "Massachusetts Avenue", "Commonwealth Avenue", "Columbus Avenue", 
  "Brighton Avenue", "North Harvard Street", "Longwood Avenue", 
  "Boston University Bridge", "Southwest Corridor"
)
busy_road_stats <- bike_network %>%
  as_tibble() %>% 
  filter(STREET_NAM %in% busy_street_names) %>%
  group_by(STREET_NAM) %>%
  summarize(total_crashes = sum(crash_count, na.rm = TRUE)) %>%
  arrange(desc(total_crashes))

ggplot(busy_road_stats, aes(x = reorder(STREET_NAM, total_crashes), y = total_crashes)) +
  geom_col(fill = "blue") +
  coord_flip() +
  labs(
    title = "Total Bike Crashes on Boston's Busiest Cycling Roads",
    subtitle = "Based on Vision Zero Crash Data and Boston Magazine's Top 10 List",
    x = "Street Name",
    y = "Number of Crashes"
  ) +
  theme_minimal()

```
```{r}
traffic_counts <- tibble(
  STREET_NAM = c("Massachusetts Avenue", "Boston University Bridge", "Commonwealth Avenue", 
                 "Southwest Corridor", "Longwood Avenue", "North Harvard Street", 
                 "Columbus Avenue", "Brighton Avenue"),
  daily_bikers = c(3878, 1957, 1571, 1554, 1348, 1320, 2139, 1063)
)

normalized_stats <- bike_network %>%
  as_tibble() %>%
  filter(STREET_NAM %in% traffic_counts$STREET_NAM) %>%
  group_by(STREET_NAM, ExisFacil) %>%
  summarize(total_crashes = sum(crash_count, na.rm = TRUE), .groups = "drop") %>%
  left_join(traffic_counts, by = "STREET_NAM") %>%
  mutate(crash_rate = (total_crashes / daily_bikers) * 100)

ggplot(normalized_stats, aes(x = reorder(STREET_NAM, crash_rate, sum), 
                             y = crash_rate, 
                             fill = ExisFacil)) +
  geom_col() +
  coord_flip() +
  scale_fill_viridis_d() +
  labs(
    title = "Bike Crash Rate on Boston's Busiest Cycling Roads",
    subtitle = "Normalized: (Total Crashes / Daily Bicyclists) * 100",
    x = "Street Name",
    y = "Total Crashes Recorded / Daily Riders",
    fill = "Bike Infrastructure Type"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.justification = "left",
    legend.box.just = "left",
    legend.text = element_text(size = 8),
    legend.title = element_text(face = "bold", size = 9)
  )
```

```{r, fig.width=10, fig.height=6, dpi=300}
normalized_stats <- normalized_stats %>%
  mutate(ExisFacil = fct_recode(ExisFacil,
    "Shared Lane Markings" = "SLM",
    "Peak Hour Bus/Bike Lane" = "BL-PEAKBUS",
    "Contraflow Bike Street" = "CFBS",
    "Hybrid Shared Lane" = "BLSL",
    "Separated/Shared Hybrid" = "SBLSL",
    "Traffic-Calmed Shared Lane" = "SLMTC",
    "Contraflow Bike Lane" = "CFBL",
    "Standard Bike Lane" = "BL",
    "Buffered Bike Lane" = "BFBL",
    "Separated/Standard Hybrid" = "SBLBL",
    "Separated Bike Lane" = "SBL",
    "Contraflow Separated Lane" = "CFSBL",
    "Natural Shared Use Path" = "SUPN",
    "Minor Shared Use Path" = "SUPM",
    "Shared Use Path" = "SUP",
    "Walkway" = "WALK",
    "Pedestrian Zone" = "PED"
  ))

ggplot(normalized_stats, aes(x = reorder(STREET_NAM, crash_rate, sum), 
                             y = crash_rate, 
                             fill = ExisFacil)) +
  geom_col() +
  coord_flip() +
  scale_fill_viridis_d() + 
  labs(
    title = "Bike Crash Rate on Busiest Boston Roads",
    subtitle = "Normalized by Number of Daily Bicyclists (Boston Magazine)",
    x = "Street Name",
    y = "Total Crashes per 100 Daily Bikers",
    fill = "Bike Infrastructure Type"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 8)) +
  guides(fill = guide_legend(ncol = 4))
```

```{r}
```