---
title: "Bike_Lane_Regression"
author: "Ben Meyer"
date: "2026-04-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(gplots)
library(gridExtra)
library(dplyr)
library(ggplot2)
```

Importing and tidying Vision Zero fatality and injury data
```{r}
vision_zero_fatality <- read_csv("C:/Users/benmi/Downloads/vision_zero_fatality.csv", 
    col_types = cols(date_time = col_datetime(format = "%Y-%m-%d %H:%M:%S +%z"), 
        mode_type = col_factor(levels = c("bike", 
            "ped", "mv")), location_type = col_factor(levels = c("Street", 
            "Intersection", "Other")))) |>
  filter(mode_type=="bike" & !is.na(street)) |>
  mutate(num_fatalities=1) |>
  group_by(street) |>
  summarize(sum(num_fatalities)) |>
  rename("num_fatalities"="sum(num_fatalities)") |>
  arrange(desc(num_fatalities)) |>
  rename(full_street_name=street)

vision_zero_crash <- read_csv("C:/Users/benmi/Downloads/vision_zero_crash.csv", 
    col_types = cols(dispatch_ts = col_datetime(format = "%Y-%m-%d %H:%M:%S+%z"), 
        mode_type = col_factor(levels = c("bike", 
            "ped", "mv")), location_type = col_factor(levels = c("Street", 
            "Intersection", "Other")))) |>
  filter(mode_type=="bike" & !is.na(street)) |>
  mutate(num_crashes=1) |>
  group_by(street) |>
  summarize(sum(num_crashes)) |>
  rename("num_crashes"="sum(num_crashes)") |>
  arrange(desc(num_crashes)) |>
  rename(full_street_name=street)
```

Creating dictionary for street names based on address data for converting abbreviations to full names, and later joining datasets with different name formatting.
```{r}
Street_Name_Dict <- read_csv("C:/Users/benmi/Downloads/SAM_Addresses.csv", show_col_types = FALSE) |>
  select(c(1:2, 13, 16, 18, 21)) |>
  janitor::clean_names() |>
  mutate(fuller_street_name = paste(street_body, street_full_suffix, sep=" ")) |>
  mutate(full_street_name = toupper(full_street_name)) |>
  select(c(3,7)) |>
  distinct()
```

Importing and tidying data about Boston streets and their corresponding bike lanes
```{r}
boston_streets <- read_csv("C:/Users/benmi/Downloads/existing_bike_network_2024.csv", 
     col_types = cols(ExisFacil = col_factor(levels = c(
#       "SBLBL", "PED", "CFSBL", "SBLSL", "BFBL", "WALK", "BLSL", "CFBS", "CFBL",
#       "SBL", "SLM", "BL", "SUP", "SUPM", "BL_PEAKBUS", "SMLTC", "SUPN")))) |>
       "SLM", "BL_PEAKBUS", "CFBS", "BLSL", "SBLSL", "SMLTC", "CFBL", "BL",
       "BFBL", "SBLBL", "SBL", "CFSBL", "SUPN", "SUPM", "SUP", "WALK", "PED"
     )))) |>
  janitor::clean_names() |>
  rename(fuller_street_name=street_nam) |>
  select(1:2, 7:8) |>
  group_by(fuller_street_name, exis_facil) |>
  summarize(install_dat=mean(install_dat))


facil_count <- existing_bike_network_2024 <- read_csv("C:/Users/benmi/Downloads/existing_bike_network_2024.csv", 
    col_types = cols_only(ExisFacil = col_guess())) |>
  janitor::clean_names() |>
  mutate(count=1) |>
  group_by(exis_facil) |>
  summarize(count=sum(count))
  

boston_bike_crashes <- boston_streets |>
  left_join(Street_Name_Dict, by=("fuller_street_name"="fuller_street_name"), relationship="many-to-many") |>
  left_join(vision_zero_crash, by=("full_street_name"="full_street_name")) |>
  left_join(vision_zero_fatality, by=("full_street_name"="full_street_name")) |>
  rename(num_injuries=num_crashes) |>
  rename(street=fuller_street_name, install_date=install_dat) |>
  filter(!is.na(street) & !is.na(full_street_name)) |>
  mutate(num_injuries = replace(num_injuries, is.na(num_injuries), 0)) |>
  mutate(num_fatalities = replace(num_fatalities, is.na(num_fatalities), 0)) |>
  mutate(sum_crashes = num_injuries+num_fatalities) |>
  mutate(sum_crashes = replace(sum_crashes, is.na(sum_crashes), 0)) |>
  select(-4)
```

Manipulating data the same way to get data about motor vehicle crashes to compare to the bike crash data
```{r}
vision_zero_fatality_mv <- read_csv("C:/Users/benmi/Downloads/vision_zero_fatality.csv", 
    col_types = cols(date_time = col_datetime(format = "%Y-%m-%d %H:%M:%S +%z"), 
        mode_type = col_factor(levels = c("bike", 
            "ped", "mv")), location_type = col_factor(levels = c("Street", 
            "Intersection", "Other")))) |>
  filter(mode_type=="mv" & !is.na(street)) |>
  mutate(num_fatalities=1) |>
  group_by(street) |>
  summarize(sum(num_fatalities)) |>
  rename("num_fatalities"="sum(num_fatalities)") |>
  arrange(desc(num_fatalities)) |>
  rename(full_street_name=street)

vision_zero_crash_mv <- read_csv("C:/Users/benmi/Downloads/vision_zero_crash.csv", 
    col_types = cols(dispatch_ts = col_datetime(format = "%Y-%m-%d %H:%M:%S+%z"), 
        mode_type = col_factor(levels = c("bike", 
            "ped", "mv")), location_type = col_factor(levels = c("Street", 
            "Intersection", "Other")))) |>
  filter(mode_type=="mv" & !is.na(street)) |>
  mutate(num_crashes=1) |>
  group_by(street) |>
  summarize(sum(num_crashes)) |>
  rename("num_crashes"="sum(num_crashes)") |>
  arrange(desc(num_crashes)) |>
  rename(full_street_name=street)

boston_mv_crashes <- boston_streets |>
  left_join(Street_Name_Dict, by=("fuller_street_name"="fuller_street_name"), relationship="many-to-many") |>
  left_join(vision_zero_crash_mv, by=("full_street_name"="full_street_name")) |>
  left_join(vision_zero_fatality_mv, by=("full_street_name"="full_street_name")) |>
  rename(num_injuries=num_crashes) |>
  rename(street=fuller_street_name, install_date=install_dat) |>
  filter(!is.na(street) & !is.na(full_street_name)) |>
  mutate(num_injuries = replace(num_injuries, is.na(num_injuries), 0)) |>
  mutate(num_fatalities = replace(num_fatalities, is.na(num_fatalities), 0)) |>
  mutate(sum_crashes = num_injuries+num_fatalities) |>
  mutate(sum_crashes = replace(sum_crashes, is.na(sum_crashes), 0)) |>
  select(-4) |>
  group_by(exis_facil) |>
  summarize(num_injuries=sum(num_injuries),
            num_fatalities=sum(num_fatalities),
            sum_crashes=sum(sum_crashes)) |>
  filter(!is.na(exis_facil)) |>
  mutate(facil_id = c(1:14)) |>
  mutate(num_fatalities=replace(num_fatalities, num_fatalities==0, NA)) |>
  left_join(facil_count, by=("exis_facil"="exis_facil")) |>
  mutate(crash_proportion_mv = sum_crashes/count) |>
  arrange(desc(crash_proportion_mv)) |>
  select(1,7)
```

Rearranging bike crash and street data to get the proportion of bike crashes per the number of the type of facility, and organizing the data
```{r}
crashes_by_facility <- boston_bike_crashes |>
  group_by(exis_facil) |>
  summarize(num_injuries=sum(num_injuries),
            num_fatalities=sum(num_fatalities),
            sum_crashes=sum(sum_crashes)) |>
  filter(!is.na(exis_facil)) |>
  mutate(facil_id = c(1:14)) |>
  mutate(num_fatalities=replace(num_fatalities, num_fatalities==0, NA)) |>
  left_join(facil_count, by=("exis_facil"="exis_facil")) |>
  mutate(crash_proportion = sum_crashes/count) |>
  arrange(desc(crash_proportion)) |>
  left_join(boston_mv_crashes, by=("exis_facil"="exis_facil"))
```

Creating a dictionary/codebook for facility types, their IDs, and what they actually represent
```{r}
facilities <- c("Separated Bike Lane on One Side, Bike Lane on Other Side",
                "Pedestrianized Street", "Contraflow Separated Bike Lane",
                "Separated Bike Lane on One Side, Shared Lane on Other Side",
                "Buffered Bus Lane", "Walkway",
                "Bike Lane on One Side, Shared Lane on Other Side", "Unclear",
                "Contraflow Bike Lane", "Separated Bike Lane",
                "Shared Lane Markings or Shared Bus Lane", "Bike Lane",
                "Shared Use Path", "Minor Shared Use Path")

facility_dict <- crashes_by_facility |>
  select(5,1) |>
  mutate(exis_facil_long=facilities)
```

Displaying the codebook
```{r}
knitr::kable(facility_dict, caption="Codebook for Bike Facility IDs, Ordered by Number of Crashes (Decreasing)")
```

Creating and displaying a linear model and graph of the proportions of crashes at each facility, by ID
```{r}
crash_model <- lm(crash_proportion~1+facil_id, data=crashes_by_facility)

ggplot(crashes_by_facility, aes(x=facil_id, y=crash_proportion)) +
  geom_point() +
  geom_smooth(method="lm", formula=y~1+x, se=FALSE) +
  labs(title="Bicycle Facility vs Proportion of Crashes", x="Bicycle Facility ID", y="Total Bike Crashes Per Facility Type Count")
```

Creating and displaying a linear model and graph of the ID
```{r}
crash_model <- lm(crash_proportion~1+facil_id, data=crashes_by_facility)

ggplot(crashes_by_facility, aes(x=facil_id, y=crash_proportion_mv)) +
  geom_point() +
  geom_point(aes(y=num_fatalities-1), color="red", size=3) +
  geom_smooth(method="lm", formula=y~1+x, se=FALSE) +
  labs(title="MV Facility vs Number of Crashes", x="Bicycle Facility ID", y="Total MV Crashes Per Facility Type Count", subtitle="red dots on facilities on which a bike fatality has occured")
```

Creating and displaying a residual and qq plot of the linear model
```{r}
residual_plot <- ggplot(crash_model, aes(crash_model$fitted.values, y=crash_model$residuals)) +
  geom_point() +
  geom_hline(yintercept=0) +
  labs(title="Residual Plot of Linear Model of Bike Crashes by Facility",
       x="Fitted Value", y="Residual")

qq_plot <- ggplot(crash_model, aes(sample=crash_model$residuals)) +
  geom_qq() +
  labs(title="Q-Q Plot of Residuals of Linear Model", x="Theoretical", y="Sample")

grid.arrange(residual_plot, qq_plot, nrow=2)
```


Creating and displaying a residual and qq plot of a second order  model
```{r}
crash_model_2 <- lm(crash_proportion~1+poly(facil_id,2), data=crashes_by_facility)

residual_plot <- ggplot(crash_model_2, aes(crash_model_2$fitted.values, y=crash_model_2$residuals)) +
  geom_point() +
  geom_hline(yintercept=0) +
  labs(title="Residual Plot of Linear Model of Bike Crashes by Facility",
       x="Fitted Value", y="Residual")

qq_plot <- ggplot(crash_model_2, aes(sample=crash_model_2$residuals)) +
  geom_qq() +
  labs(title="Q-Q Plot of Residuals of Linear Model", x="Theoretical", y="Sample")

grid.arrange(residual_plot, qq_plot, nrow=2)
```