---
title: "Data Science Final Project"
author: "Amanda Chang, Ben Meyer, Trinity Lee"
output: html_notebook
---



```{r}
library(tidyverse)
library(lubridate)
library(leaflet)
library(shiny)
library(leaflet)
library(dplyr)
library(readr)
library(lubridate)
library(glue)
```

Load Bike Accidents Data & Add Seasons 
```{r}
data <- read_csv("Bike_Crash_Data_Vision_Zero.csv")

bike <- data |>
  filter(mode_type == "bike") |>
  mutate(
    dispatch_ts = ymd_hms(dispatch_ts),
    date = as_date(dispatch_ts),
    year = year(dispatch_ts),
    season = case_when(
            month(date) %in% c(12, 1, 2) ~ "Winter",
            month(date) %in% c(3, 4, 5) ~ "Spring",
            month(date) %in% c(6, 7, 8) ~ "Summer",
            month(date) %in% c(9, 10, 11) ~ "Fall"
          ),
    time_numeric = hour(dispatch_ts) )

```

Shiny Slider of Bike Accidents
```{r}
ui <- fluidPage(
  titlePanel("Bike Crashes by Time of Day"),
  
  sidebarLayout(sidebarPanel(sliderInput(
        "time",
        "Show crashes at hours:",
        min = 0,
        max = 24,
        value = 12,
        step = 1
      )),
    mainPanel(
      leafletOutput("map")
    )))

server <- function(input, output, session) {
  filtered_data <- reactive({
    bike |>
      filter(time_numeric == input$time)
  })
  
  # Base map
  output$map <- renderLeaflet({
    leaflet(bike) |>
      addTiles()  |>
      setView(lng = mean(bike$long, na.rm = TRUE),
              lat = mean(bike$lat, na.rm = TRUE),
              zoom = 12)
  })
  
  # Update points when slider moves
  observe({
    leafletProxy("map", data = filtered_data()) |>
      clearMarkers() |>
      addCircleMarkers(
        lng = ~long,
        lat = ~lat,
        color = "blue",
        radius = 4,
        stroke = FALSE,
        fillOpacity = 0.7,
        popup = ~paste(
          "Year:", year, "<br>",
          "Mode:", mode_type, "<br>",
          "Street:", street, "<br>",
          "Time:", dispatch_ts
        ))})
}
shinyApp(ui, server)
```

Bar charts of Accidents 

```{r}
ggplot(bike, aes(x = time_numeric, color=season)) +
  geom_histogram(binwidth=1) +
  facet_wrap(season ~.)+
  labs(
    title = "Distribution of Crash Times",
    x = "Hour of Day",
    y = "Number of Crashes"
  ) +
  theme_minimal()

ggplot(bike, aes(x = time_numeric)) +
  geom_histogram(binwidth=1) +
  facet_wrap(year ~.)+
  labs(
    title = "Distribution of Crash Times",
    x = "Hour of Day",
    y = "Number of Crashes"
  ) +
  theme_minimal()
```

Processing Blue Bikes Data
```{r}
#years <- list("2015", "2016", "2017", "2018", "2019", "2020" , "2021", "2022", "2023", "2024", "2025")     
# Mass processing data

#new_data <- data.frame()
#for (month in months) {
#  filename <- glue("blue_bikes_data/{year}_data/{year}{month}-bluebikes-tripdata/{year}{month}-bluebikes-tripdata.csv")
#  temp_data <- read_csv(filename, col_types = cols(.default = col_character()))
#  blue_data <- bind_rows(blue_data, temp_data)
#}
#write.csv(blue_data, glue("blue_bikes_data/{year}_data.csv"))


#for (year in years) {
#  filename <- glue("blue_bikes_data_trimmed/{year}_data.csv")
#  temp_data <- read_csv(filename, col_types = cols(.default = col_character()))
#  new_data <- temp_data |>
#  select(
#    starttime,
#    stoptime,
#    `start station id`,
#    `start station name`,
#    `start station latitude`,
#    `start station longitude`,
#    `end station id`,
#    `end station name`,
#    `end station latitude`,
#    `end station longitude`
#  )
#  write.csv(new_data, glue("blue_bikes_data_trimmed/trimmed_{year}_data.csv"))
#}
#for (year in list("2024", "2025")) {
#  filename <- glue("blue_bikes_data_trimmed/{year}_data.csv")
#  temp_data <- read_csv(filename, col_types = cols(.default = col_character()))
#new_data <- temp_data |>
#  rename(
#    starttime = started_at,
#    stoptime = ended_at,
#    `start station id` = start_station_id,
#    `start station name` = start_station_name,
#    `start station latitude` = start_lat,
#    `start station longitude` = start_lng,
#    `end station id` = end_station_id,
#    `end station name` = end_station_name,
#    `end station latitude` = end_lat,
#   `end station longitude` = end_lng
# ) |>
# select(
#    starttime,
#    stoptime,
#    `start station id`,
#    `start station name`,
#    `start station latitude`,
#    `start station longitude`,
#    `end station id`,
#    `end station name`,
#    `end station latitude`,
#    `end station longitude`
#  )
#  write.csv(new_data, glue("blue_bikes_data_trimmed/trimmed_{year}_data.csv"))
#}

```

Group Blue Bikes based on Station Name & Latitude/Longitude and Find Total Trips Per Station
```{r}
#years <- list("2015", "2016", "2017", "2018", "2019", "2020" , "2021", "2022", "2023", "2024", "2025")     
#total_counts <- data.frame(
#  start_station = character(),
#  start_lat = numeric(),
#  start_lng = numeric(),
#  n = numeric()
#)
#
#for (year in years) {
#  df <- read_csv(glue("blue_bikes_data_start_trimmed/start_trimmed_{year}_data.csv"))
#  
#  temp <- df |>
#    group_by(start_station, start_lat, start_lng) |>
#    summarise(n = n(), .groups = "drop")
#  
#  total_counts <- bind_rows(total_counts, temp)
#}
#
#total_counts <- total_counts |>
#  group_by(start_station) |>
#  summarise(
#    n = sum(n),  # ← sum counts across years
#    start_lat = mean(start_lat, na.rm = TRUE),
#    start_lng = mean(start_lng, na.rm = TRUE),
#    .groups = "drop"
#  )
#
#
```
  
  
Import Total Trip Count Per Start Station with Coordinations and Filter Out Stations Outside of Boston

```{r}

total_counts <- read.csv("total_counts_with_coord.csv")
total_counts <- head(total_counts, -1) |>
  #only in boston
  filter((42.237210 < start_lat & start_lat < 42.400739) & 
         ( -71.186470 < start_lng & start_lng < -70.985085)) |>
  
  #cambridge
  filter(!(
    start_lat > 42.355417 & start_lat < 42.425723 &
    start_lng > -71.114138 & start_lng < -71.088989
  )) |>
  
  #somerville
  #42.374595, -71.168457
  #42.432250, -71.098786
  filter(!(
    start_lat > 42.374595 & start_lat < 42.432250 &
    start_lng > -71.168457 & start_lng < -71.098786
  )) |>

  #harvard
  # 42.370632, -71.124589
  # 42.377532, -71.115612
  filter(!(
    start_lat > 42.370632 & start_lat < 42.377532 &
    start_lng >  -71.124589 & start_lng < -71.115612
  )) |>

  #42.364358, -71.117201
  #42.371603, -71.113690
  filter(!(
    start_lat > 42.364358 & start_lat <42.371603 &
    start_lng > -71.117201 & start_lng < -71.113690
  )) |>
  
  #42.355546, -71.094754
  #42.380322, -71.079337
  filter(!(
    start_lat >42.355546 & start_lat < 42.380322 &
    start_lng > -71.094754 & start_lng <  -71.079337
  )) 

total_counts <- total_counts |>
  filter(!is.na(start_lat), !is.na(start_lng), !is.na(n))
```



General Heatmap based on Number of Trips Per Station
```{r}
pal <- colorNumeric(
  palette = "YlOrRd",
  domain = total_counts$n
)

leaflet(total_counts) |>
  addTiles() |>
  addCircleMarkers(
    lat = ~start_lat,
    lng = ~start_lng,
    color = ~pal(n),
    fillColor = ~pal(n),
    fillOpacity = 0.8,
    stroke = TRUE,
    weight = 1,
    radius = 7,
    popup = ~paste0(
      "<b>", start_station, "</b><br/>",
      "Count: ", n
    ),
    label = ~start_station
  )|>
  addLegend(
    position = "bottomright",
    pal = pal,
    values = ~n,
    title = "Number of Trips Per Station",
    opacity = 0.8
    )
```


Bike Accidents Overlayed onto Station Trips
```{r}
bike <- bike |>
  filter(!is.na(lat), !is.na(long))

pal <- colorNumeric(
  palette = "YlOrRd",
  domain = total_counts$n
)

#Station Data
leaflet()|>
  addTiles() |>
  addCircles(
    data = total_counts,
    lat = ~start_lat,
    lng = ~start_lng,
    radius = ~sqrt(n)/3, #Stations with More Trips Are darler & have larger radius
    fillColor = ~pal(n),
    stroke= FALSE,
    fillOpacity = 0.5,   # lower = more blending
  ) |>
  
  #Accident Data
    addCircleMarkers(
    data = bike,
    lat = ~lat,
    lng = ~long,
    radius = 2,
    color = "blue",
    stroke= FALSE,
    fillOpacity = 0.7,
  ) |>
  
  addLegend(
    position = "bottomright",
    pal = pal,
    values = total_counts$n,
    title = "Number of Blue Bike Trips Per Station",
    opacity = 0.8
  )

```


Grid View of Trips to Accident Ration Based on Latitude/Longitude
```{r}
cell_size <- 0.002 #Grid Cell Size  

acc_grid <- bike |>
  mutate(
    lat_bin = floor(lat / cell_size) * cell_size,
    lng_bin = floor(long / cell_size) * cell_size
  )

bike_grid <- total_counts |>
  mutate(
    lat_bin = floor(start_lat / cell_size) * cell_size,
    lng_bin = floor(start_lng / cell_size) * cell_size
  )

# Count accidents per grid cell
acc_counts <- acc_grid |>
  group_by(lat_bin, lng_bin) |>
  summarize(bike = n(), .groups = "drop")

# Count Bike Trips Per Grid
bike_counts <- bike_grid |>
  group_by(lat_bin, lng_bin) |>
  summarize(trips = sum(n), .groups = "drop")  # ← Changed from n() to sum()

grid_data <- full_join(acc_counts, bike_counts,
                       by = c("lat_bin", "lng_bin")) |>
  mutate(
    bike = ifelse(is.na(bike), 0, bike),
    trips = ifelse(is.na(trips), 0, trips),
    rate = ifelse(trips == 0, 0, bike / trips)
  )

#Take out squares with only 1 accident
grid_data <- grid_data |>
  filter(bike > 1)

mean(grid_data$bike)
```


```{r}
#Grid Heatmap Based on Accident to Trips Ratio
pal <- colorNumeric(
  palette = "YlOrRd",  # Yellow-Orange-Red
  domain = grid_data$rate,
)
leaflet(grid_data)|>
  addTiles() |>
  addRectangles(
    lng1 = ~lng_bin,
    lat1 = ~lat_bin,
    lng2 = ~lng_bin + cell_size,
    lat2 = ~lat_bin + cell_size,
    fillColor = ~pal(rate),
    fillOpacity = 0.7,
    color = NA,
    popup = ~paste0(
      "Accidents: ", bike, "<br>",
      "Trips: ", trips, "<br>",
      "Accidents to Trip Ratio: ", round(rate, 5)
    )
  ) |>
  addLegend(
    pal = pal,
    values = ~rate,
    title = "Accidents to Total Trips Ratio"
  )

#Grid Heatmap Based on Bike Accidents Only

cap_value <- quantile(grid_data$bike, 0.95) # Cap at 95th percentile
pal <- colorNumeric(
  palette = "YlOrRd",
  domain = c(0, cap_value),
  na.color = "transparent"
)

leaflet(grid_data)|>
  addTiles() |>
  addRectangles(
    lng1 = ~lng_bin,
    lat1 = ~lat_bin,
    lng2 = ~lng_bin + cell_size,
    lat2 = ~lat_bin + cell_size,
    fillColor = ~pal(pmin(bike, cap_value)),  
    fillOpacity = 0.7,
    color = NA,
    popup = ~paste0(
      "Accidents: ", bike, "<br>",
      "Trips: ", trips, "<br>",
      "Accidents to Trip Ratio: ", round(rate, 5)
    )
  ) |>
    addLegend(
    pal = pal,
    values = ~pmin(bike, cap_value),
    opacity = 0.7,
    title = "Accidents per Station(capped at 11+)"

  )
```



Merging Amanda's Bike Lanes with My Maps



```{r, include=FALSE}

library(DT)            # for visualizing data tables 
library(janitor)       # for automatically renaming variables
library(ggmap)         # for mapping points on maps
library(RColorBrewer)  # for color palettes
library(sf)            # for working with spatial data
library(gplots)        # for col2hex() function
library(choroplethr)   # another option to make choropleths
library(choroplethrMaps)
library(maps)

boston_crashes_raw <- bike

bike_network <- st_read("existing_bike_network_2024.json") %>% 
  st_transform(4326) # Ensure WGS84 for Leaflet
head(bike_network)

bike_network_2022 <- st_read("existing_bike_network_2022.json") %>% 
  st_transform(4326) # Ensure WGS84 for Leaflet
head(bike_network_2022)
```

```{r}
unique(bike_network$ExisFacil)
```

```{r}
bike_crashes <- boston_crashes_raw %>%
  filter(mode_type == "bike") %>%
  drop_na(lat, long) %>%
  st_as_sf(coords = c("long", "lat"), crs = 4326)

roads_projected <- st_transform(bike_network, 2249)
crashes_projected <- st_transform(bike_crashes, 2249)

nearest_road_indices <- st_nearest_feature(crashes_projected, roads_projected)

road_counts <- table(nearest_road_indices)
bike_network$crash_count <- 0
bike_network$crash_count[as.numeric(names(road_counts))] <- as.numeric(road_counts)

# sorted from most dangerous (top of legend) to safest (bottom of legend)
safety_order <- c(
  "SLM",        # Shared Lane Markings (Zero protection, full mixing)
  "BL-PEAKBUS", # Bike Lane/Peak Hour Bus Lane (High-speed heavy vehicle mixing)
  "CFBS",       # Contraflow Bike Street (Shared space with opposing traffic)
  "BLSL",       # Bike Lane/Shared Lane Hybrid (Inconsistent protection)
  "SBLSL",      # Separated Bike Lane/Shared Lane Hybrid (Gaps in protection)
  "SLMTC",      # Traffic-Calmed Local Street with Shared Lane Markings
  "CFBL",       # Contraflow Bike Lane (Painted-only, high conflict risk)
  "BL",         # Bike Lane (Standard painted lane, door-zone risk)
  "BFBL",       # Buffered Bike Lane (Paint buffer, but no physical barrier)
  "SBLBL",      # Separated Bike Lane/Bike Lane Hybrid
  "SBL",        # Separated Bike Lane (Standard flex-post or curb)
  "CFSBL",      # Contraflow Separated Bike Lane (High protection)
  "SUPN",       # Shared Use Path - Natural Surface (Safe from cars, but slip risk)
  "SUPM",       # Shared Use Path - Minor
  "SUP",        # Shared Use Path (Fully off-street)
  "WALK",       # Walkway
  "PED"         # Pedestrian Zone (Lowest vehicle interaction)
)

bike_network$ExisFacil <- factor(bike_network$ExisFacil, levels = safety_order)


```


```{r}
pal1 <- colorFactor(
  palette = c("#e41a1c", "#6250FF", "#00A719"),
  domain = bike_network$ExisFacil
)

pal2 <- colorNumeric(
  palette = "YlOrRd",  # Yellow-Orange-Red
  domain = grid_data$bike,
)

leaflet() %>%
  
    addRectangles(
    data = grid_data,
    lng1 = ~lng_bin,
    lat1 = ~lat_bin,
    lng2 = ~lng_bin + cell_size,
    lat2 = ~lat_bin + cell_size,
    fillColor = ~pal2(bike),
    fillOpacity = 0.7,
    color = NA,
    group = "Crash Hotspots Ratio",
    popup = ~paste0(
      "Accidents: ", bike, "<br>",
      "Trips: ", trips, "<br>",
      "Accidents to Trip Ratio: ", round(rate, 5)
    )
  ) %>%
  
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolylines(
    data = bike_network,
    color = ~pal1(ExisFacil),
    weight = 3,
    opacity = 1,
    label = ~paste0("Street: ", STREET_NAM, " | Crashes: ", crash_count),
    group = "Infrastructure"
  ) %>%

  
  addLayersControl(
    overlayGroups = c("Infrastructure", "Crash Hotspots Ratio"),
    options = layersControlOptions(collapsed = FALSE)
  ) %>%
  
  addLegend(
    pal = pal,
    values = bike_network$ExisFacil,
    title = "Road Category",
    position = "bottomright"
  )

```

