---
title: "Crashes By Facility Type Tidying"
author: "Amanda Chang, Ben Meyer, Trinity Lee"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
library(readr)
library(dplyr)
library(tidyverse)
library(janitor)
```

# Tidying and Joining Data for Bike Facility Regression
## Vision Zero Data (Bike Crashes)
Function to tidy vision zero data and isolate counts of crashes at each street
```{r}
tidy_vision_zero_data <- function(dir) {
  street_data <- read_csv(dir, 
    col_types = cols(location_type = col_factor(levels = c("Street", 
            "Intersection", "Other")))) |>
    filter(mode_type=="bike", location_type=="Street") |>
    select(c(street, location_type))
  
  intersection_data <- read_csv(dir, 
    col_types = cols(location_type = col_factor(levels = c("Street", 
            "Intersection", "Other")))) |>
    filter(mode_type=="bike", location_type=="Intersection") |>
    select(!street) |>
    pivot_longer(cols=c(xstreet1, xstreet2), names_to="street_type", values_to="street") |>
    select(c(street, location_type))
  
  total_data <- rbind(street_data, intersection_data) |>
    group_by(street) |>
    summarize(count=n()) |>
    arrange(desc(count))
  
  return(total_data)
}
```

Function to swap between street abbreviations and full names
```{r}
roadway_titles = tibble(
  abbreviation = c("HWY", "RD", "ST", "AVE", "WHRF", "DR", "BRG", "SQ", "PKWY", "BLVD", "TER", "PLZ"),
  full = c("HIGHWAY", "ROAD", "STREET", "AVENUE", "WHARF", "DRIVE", "BRIDGE", "SQUARE", "PARKWAY", "BOULEVARD", "TERRACE", 
           "PLAZA")
)

convert_roadnames <- function(name) {
  for(i in 1:length(roadway_titles$full)) {
    name <- gsub(roadway_titles$abbreviation[i], roadway_titles$full[i], name)
  }
  return(name)
}
```

Importing and tidying Vision Zero fatality and injury data
```{r}
vision_zero_fatality <- tidy_vision_zero_data("../data/Bike_Fatality_Data_Vision_Zero.csv")
vision_zero_injury <- tidy_vision_zero_data("../data/Bike_Crash_Data_Vision_Zero.csv")
vision_zero_total <- full_join(vision_zero_injury, vision_zero_fatality, by=("street"="street")) |>
  rename(injury_count=`count.x`, fatality_count=`count.y`) |>
  mutate(fatality_count=replace(fatality_count, is.na(fatality_count), 0),
         sum_crashes=injury_count+fatality_count) |>
  arrange(desc(sum_crashes)) |>
  mutate(streetname=convert_roadnames(street)) |>
  select(streetname, injury_count, fatality_count, sum_crashes)
rm(vision_zero_fatality, vision_zero_injury)
```

## Boston Managed Streets (Data About Streets)
Getting street lengths from Boston Gov Data
```{r}
boston_streets <- read_csv("../data/City_Of_Boston_Managed_Streets.csv", show_col_types=FALSE) |>
  clean_names() |>
  filter(mgis_town=="BOSTON", !is.na(streetname)) |>
  select(c(streetname, length_ft, num_lanes)) |>
  group_by(streetname) |>
  summarize(length_ft=sum(length_ft), num_lanes=mean(num_lanes)) |>
  mutate(area=length_ft*num_lanes)
```

## Boston Bike Network (Bike Facilities by Street)
Getting bike facilities from Boston Bike Network Data
```{r}
bike_network <- read_csv("../data/existing_bike_network_2024.csv", 
     col_types = cols(ExisFacil = col_factor(levels = c(
#       "SBLBL", "PED", "CFSBL", "SBLSL", "BFBL", "WALK", "BLSL", "CFBS", "CFBL",
#       "SBL", "SLM", "BL", "SUP", "SUPM", "BL-PEAKBUS", "SLMTC", "SUPN"
       "SLM", "BL-PEAKBUS", "CFBS", "BLSL", "SBLSL", "SLMTC", "CFBL", "BL",
       "BFBL", "SBLBL", "SBL", "CFSBL", "SUPN", "SUPM", "SUP", "WALK", "PED"
     )))) |>
  janitor::clean_names() |>
  select(c(street_nam, exis_facil)) |>
  rename(streetname=street_nam) |>
  group_by(streetname, exis_facil) |>
  summarize() |>
  mutate(streetname=toupper(streetname))
```

## Joining Data
Joining together Vision Zero and Boston Bike Network data
```{r}
crashes_by_street <- vision_zero_total |>
  left_join(bike_network, by=("streetname"="streetname")) |>
  left_join(boston_streets, by=("streetname"="streetname")) |>
  mutate(crashes_per_distance=sum_crashes/length_ft,
         crashes_per_lane=sum_crashes/num_lanes,
         crashes_per_area=sum_crashes/area) |>
  arrange(desc(crashes_per_area)) |>
  filter(num_lanes>0)
```

Grouping crashes by bike facility type
```{r}
crashes_by_facility <- crashes_by_street |>
  filter(!is.na(exis_facil)) |>
  mutate(exis_facil = replace(exis_facil, exis_facil=="CFBS", "CFBL")) |>
  group_by(exis_facil) |>
  summarize(crashes_per_area=mean(crashes_per_area),
            crashes_per_lane=mean(crashes_per_lane),
            crashes_per_distance=mean(crashes_per_distance),
            sum_crashes=mean(sum_crashes),
            num_facil=n()) |>
  arrange(desc(crashes_per_area))

crashes_by_facility$facil_id <- 1:14
```

Creating codebook to show and explain facility types
```{r}
facilities <- c("Separated Bike Lane on One Side, Shared Lane on Other Side",
                "Separated Bike Lane on One Side, Bike Lane on Other Side",
                "Bike Lane", "Walkway", "Separated Bike Lane",
                "Contraflow Separated Bike Lane",
                "Bike Lane on One Side, Shared Lane on Other Side",
                "Shared Lane Markings or Shared Bus Lane", 
                "Bike Lane / Peak Hour Bus Lane",
                "Buffered Bus Lane", "Pedestrianized Street",
                "Contraflow Bike Lane", "Shared Use Path",
                "Shared Lane Markings on a Traffic Calmed Street")

facility_codebook <- crashes_by_facility |>
  select(7,1) |>
  mutate(exis_facil_long=facilities)
```

Creating table with proportions of each facility type
```{r}
facility_counts <- crashes_by_facility |>
  select(c(1, 2, 6)) |>
  #arrange(desc(num_facil)) |>
  mutate(proportion=num_facil/sum(num_facil)) |>
  left_join(facility_codebook, by=("exis_facil"="exis_facil")) |>
  select(-5)
```

## Saving Cleaned/Joined Data
Writing tibbles to csv files in repository
```{r}
write_csv(facility_codebook, "../data/facility_codebook.csv")
write_csv(crashes_by_facility, "../data/crashes_by_facility.csv")
write_csv(facility_counts, "../data/facility_counts.csv")
```